---
title: "Intertidal Plot Figures"
format: pdf
editor: visual
author: "Ross Whippo"
date: today
---

```{r}
#| echo: false
#| output: false

library(tidyverse)
library(viridis)
library(ggrepel)
library(vegan)

all_data <- read_csv("data/all_data_QAQC.csv")

```

## Spatial Trends

Visualizations of patterns seen in organismal abundance across years. Initial figures consider two groupings for each organism; a group based on triplets and a group based on blocks (A = triplets 1:6, B = triplets 7:12).

{{< pagebreak >}}

### Fucus

#### Triplet Groups

```{r, fig.height=6}
#| echo: false
#| fig-cap: "Fucus cover for each triplet by year. Lines are third degree polynomial splines with 95% confidence interval."

all_data %>%
  mutate(fucus_percent_total = as.numeric(fucus_percent_total)) %>%
  mutate(triplet = factor(triplet, levels = c(1:12))) %>%
  ggplot(aes(x = year, y = fucus_percent_total, fill = triplet,
             color = triplet)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3)) +
  scale_color_viridis(discrete = TRUE) +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  facet_wrap(.~triplet, ncol = 3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

{{< pagebreak >}}

#### Blocks

```{r}
#| echo: false
#| fig-cap: "Fucus cover for blocks A (triplets 1:6) and B (triplets 7:12). Lines are third degree polynomial splines with 95% confidence interval."

all_data %>%
  mutate(fucus_percent_total = as.numeric(fucus_percent_total)) %>%
  mutate(triplet = factor(triplet, levels = c(1:12))) %>%
  mutate(block = case_when(triplet %in% c(1:6) ~ 'A',
                           triplet %in% c(7:12) ~ 'B')) %>%
  ggplot(aes(x = year, y = fucus_percent_total, fill = block,
             color = block)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3)) +
  scale_color_viridis(discrete = TRUE, option = 'magma', 
                      begin = 0.4, end = 0.7) +
  scale_fill_viridis(discrete = TRUE, option = 'magma', 
                     begin = 0.4, end = 0.7) +
  theme_minimal() +
  facet_wrap(.~block) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

{{< pagebreak >}}

### Mussels

#### Triplet Groups

```{r, fig.height=6}
#| echo: false
#| fig-cap: "Mussel cover for each triplet by year. Lines are third degree polynomial splines with 95% confidence interval."

all_data %>%
   mutate(triplet = factor(triplet, levels = c(1:12))) %>%
  ggplot(aes(x = year, y = mytilus, fill = triplet,
             color = triplet)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3)) +
  scale_color_viridis(discrete = TRUE) +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  facet_wrap(.~triplet, ncol = 3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
{{< pagebreak >}}

#### Blocks

```{r}
#| echo: false
#| fig-cap: "Mussel cover for blocks A (triplets 1:6) and B (triplets 7:12). Lines are third degree polynomial splines with 95% confidence interval."

all_data %>%
 mutate(triplet = factor(triplet, levels = c(1:12))) %>%
  mutate(block = case_when(triplet %in% c(1:6) ~ 'A',
                           triplet %in% c(7:12) ~ 'B')) %>%
  ggplot(aes(x = year, y = mytilus, fill = block,
             color = block)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x)) +
  scale_color_viridis(discrete = TRUE, option = 'magma', 
                      begin = 0.4, end = 0.7) +
  scale_fill_viridis(discrete = TRUE, option = 'magma', 
                     begin = 0.4, end = 0.7) +
  theme_minimal() +
  facet_wrap(.~block) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

{{< pagebreak >}}

## Multivariate Path

These are nMDS plots of community composition labeled by year to determine any patterns through time. Points represent the multivariate mean of the total community and lines join each year sequentially. 

### Entire Community

```{r}
#| echo: false
#| output: false

alldata_summary <- all_data %>%
  relocate(year, .before = fucus_percent_total) %>%
  group_by(year) %>%
  select(!(c(fucus_number_adults:fucus_sporelings_number,
           antho_artemesia,
           katharina,
           spirorbidae,
           petrocelia,
           ralfsia_hild,
           petrocelis))) %>%
  mutate(across(fucus_percent_total:idotea, as.numeric)) %>%
  summarise(across(fucus_percent_total:idotea, ~mean(.x)))

yearallMDS <- metaMDS(alldata_summary[,2:59], distance = "altGower")


# extract the 'points' from the nMDS that you will plot in ggplot2
yearall_points <- yearallMDS$points
# turn those plot points into a dataframe that ggplot2 can read
yearall_points <- data.frame(yearall_points)
# join your plot points with your summed species observations from each habitat type
plot_data_tax <- data.frame(yearall_points, alldata_summary[,1])

```

```{r}
#| echo: false
#| fig-cap: "Mean multidimensional community composition of across all plots per year."

# run the ggplot
  ggplot(plot_data_tax, aes(x=MDS1, y=MDS2, color = year)) +
  labs(x = "nMDS1", y = "nMDS2") +
  theme_classic() + 
  geom_point(size =  4) + 
  scale_color_viridis(discrete = FALSE, option = "G", begin = 0.1, end = 0.9, guide = "none") +
    geom_text_repel(aes(x = MDS1, y = MDS2, label = year, color = year)) +
    geom_path(arrow = arrow(angle = 15, ends = "last", type = "closed")) 
    
  
```

{{< pagebreak >}}

### Community By Block

```{r}
#| echo: false
#| output: false

alldata_summary <- all_data %>%
  relocate(year, .before = fucus_percent_total) %>%
  mutate(block = case_when(triplet %in% 1:6 ~ "A",
                           triplet %in% 7:12 ~ "B"),
         .before = year) %>%
  group_by(year, block) %>%
  select(!(c(fucus_number_adults:fucus_sporelings_number,
           antho_artemesia,
           katharina,
           spirorbidae,
           petrocelia,
           ralfsia_hild,
           petrocelis))) %>%
  mutate(across(fucus_percent_total:idotea, as.numeric)) %>%
  summarise(across(fucus_percent_total:idotea, ~mean(.x)))

yearallMDS <- metaMDS(alldata_summary[,3:60], distance = "altGower")


# extract the 'points' from the nMDS that you will plot in ggplot2
yearall_points <- yearallMDS$points
# turn those plot points into a dataframe that ggplot2 can read
yearall_points <- data.frame(yearall_points)
# join your plot points with your summed species observations from each habitat type
plot_data_tax <- data.frame(yearall_points, alldata_summary[,1:2])

```

```{r, fig.height=6}
#| echo: false
#| fig-cap: "Multidimensional mean community composition of blocks A (triplets 1:6) and B (triplets 7:12) per year."

# run the ggplot
  ggplot(plot_data_tax, aes(x=MDS1, y=MDS2, 
                            color = block, label = year)) + 
  labs(x = "nMDS1", y = "nMDS2") +
  theme_classic() + 
  geom_point(size =  3) + 
  scale_color_viridis(discrete = TRUE, begin = 0.4, end = 0.7, option = "G", name = "block") +
     geom_text_repel(aes(x = MDS1, y = MDS2, label = year, color = block), size = 4) +
    geom_path(arrow = arrow(angle = 5, ends = "last", type = "closed")) +
    facet_wrap(.~block, ncol = 1)

```

